---
title: 'HKS SUP-135 Lab 4: The Tennessee STAR Experiment'
author: "Matt Khinda"
date: "2/25/2023"
output: pdf_document
---

```{r setup, include=FALSE}
# Settings for PDF output
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_chunk$set(fig.width=12, fig.height=5.5, fig.align = "center") 

# Require packages and install and load if not already
if (!require(haven)) install.packages("haven"); library(haven)
if (!require(tidyverse)) install.packages("tidyverse"); library(tidyverse)
if (!require(ggplot2)) install.packages("ggplot2"); library(ggplot2)
if (!require(statar)) install.packages("statar"); library(statar)
if (!require(here)) install.packages("here"); library(here)
if (!require(sandwich)) install.packages("sandwich"); library(sandwich)
if (!require(lmtest)) install.packages("lmtest"); library(lmtest)

# Clear workspace
rm(list=ls())
cat('\014')
```

```{r laod data, include=FALSE}
star <- read_dta(here("datasets", "star.dta"))
```

## Question 1: The Need for Complete Randomization

In the case of the Tennessee STAR Experiment, it is important that teachers be randomly assigned to the control or treatment group (along with students) because not doing so could introduce selection bias. This might look like teachers deciding to teach smaller or larger classes based on a pre-existing preference and knowledge that they teach better under certain classroom conditions. This motivation would act as a confounding variable and obfuscate the validity of the results because in that case it may not be the class size itself that has a causal effect on outcomes, but rather (in part or in whole) the teachers' comfort with the class size.


## Question 2: Average class sizes

```{r class_size}
avg_small_class <- mean(star$class_size[star$small == 1])
avg_reg_class <- mean(star$class_size[star$small == 0])

cat("The average class size for small kindergarten classes is", avg_small_class, "
    while for regular kindergarten classes it is", avg_reg_class)
```

## Question 3: SAT Score Index

#### 3a: Standardized Scores
```{r sat scores}
ctrl_grp <- subset(star, small == 0)

math_ctrl_mean <- mean(ctrl_grp$math, na.rm = T)
math_ctrl_sd <- sd(ctrl_grp$math, na.rm = T)

read_ctrl_mean <- mean(ctrl_grp$read, na.rm = T)
read_ctrl_sd <- sd(ctrl_grp$read, na.rm = T)

word_ctrl_mean <- mean(ctrl_grp$wordskill, na.rm = T)
word_ctrl_sd <- sd(ctrl_grp$wordskill, na.rm = T)

listen_ctrl_mean <- mean(ctrl_grp$listen, na.rm = T)
listen_ctrl_sd <- sd(ctrl_grp$listen, na.rm = T)

star$std_math <- (star$math - math_ctrl_mean)/math_ctrl_sd
star$std_read <- (star$read - read_ctrl_mean)/read_ctrl_sd
star$std_word <- (star$wordskill - word_ctrl_mean)/word_ctrl_sd
star$std_listen <- (star$listen - listen_ctrl_mean)/listen_ctrl_sd
```
```{r show1, echo = FALSE}
paste("Standardized Math Score")
summary(star$std_math)

paste("Standardized Reading Score")
summary(star$std_read)

paste("Standardized Wordskill Score")
summary(star$std_word)

paste("Standardized Listening Score")
summary(star$std_listen)
```



#### 3b: Generating SAT Index
```{r sat index, message=FALSE, warning=FALSE}
# Calculate composite score based on mean of standardized scores
for (x in 1:nrow(star)){
  star$sat_index[x] <- mean(c(star$std_math[x], 
                              star$std_read[x], 
                              star$std_word[x], 
                              star$std_listen[x]), 
                            na.rm = T)}
```
```{r show2, echo = FALSE}
paste("SAT Index Score")
summary(star$sat_index)
```

#### 3c: Histograms
```{r sat hist}
ggplot() +
  geom_histogram(data = filter(star, small == 0), aes(x = sat_index), bins = 30) + 
  labs(title="Distribution of SAT Index Scores for Control Group", y = "Count", x = "Score") +
  scale_x_continuous(breaks = seq(-3,5, by =1)) +
  theme(plot.title = element_text(size=21))

ggplot() +
  geom_histogram(data = filter(star, small == 1), aes(x = sat_index), bins = 30) + 
  labs(title="Distribution of SAT Index Scores for Treatment Group", y = "Count", x = "Score") +
  scale_x_continuous(breaks = seq(-3,5, by =1)) +
  theme(plot.title = element_text(size=21))
```

As we can see in the histograms above, the distribution of SAT Index scores for both the control and treatment groups is fairly normal. That said, in the treatment group we observe a slightly stronger rightward skew which represents a higher number of overachievers (people who scored 0.5 or more standard deviations above the control group mean). When considering this, it is also worth keeping in mind that there is a notable difference in sample sizes between the control group (n = 3987) and the treatment group (n = 1723). 


## Question 4: Teacher Randomization

#### 4a: Aggregate Data by Teacher ID
```{r teacher id}
star_teachers <- star %>%
  group_by(teacher_id) %>%
  summarise(school_id = mean(school_id),
            small = mean(small),
            teacher_masters = mean(teacher_masters),
            teacher_white = mean(teacher_white),
            teacher_black = mean(teacher_black),
            teacher_experience = mean(teacher_experience),
            sat_index = mean(sat_index),
            class_size = mean(class_size))
```
```{r show3, echo = FALSE}
paste("Teacher Means in Control Group")
summary(filter(select(star_teachers, !c("teacher_id","school_id")), small == 0))
paste("Teacher Means in Treatment Group")
summary(filter(select(star_teachers, !c("teacher_id","school_id")), small == 1))
```

#### 4b & c: Measuring Differences in Predetermined Characteristics
```{r teacher diffs}
# Create linear model to test relationship between characteristic and group status
model_exp <- lm(teacher_experience ~ small, data = star_teachers)

# Get lower and upper bounds for 95% confidence interval
bounds_exp <- c(-0.35709 - 1.96*0.66372, -0.35709 + 1.96*0.66372)
cat("The estimated difference in experience between teachers in the control 
    and treatment groups is", model_exp$coefficients["small"])

model_edu <- lm(teacher_masters ~ small, data = star_teachers)
bounds_edu <- c(-0.042648  - 1.96*0.054005, -0.042648  + 1.96*0.054005)
cat("The estimated difference in the fraction of teachers with masters degrees 
     between the control and treatment groups is", model_edu$coefficients["small"])

model_white <- lm(teacher_white ~ small, data = star_teachers)
bounds_white <- c(0.043565  - 1.96*0.040778, 0.043565 + 1.96*0.040778)
cat("The estimated difference in the fraction of white teachers 
     between the control and treatment groups is", model_white$coefficients["small"])


model_blk <- lm(teacher_black ~ small, data = star_teachers)
bounds_blk <- c(-0.038489  - 1.96*0.040571, -0.038489 + 1.96*0.040571)
cat("The estimated difference in the fraction of white teachers 
     between the control and treatment groups is", model_blk$coefficients["small"])
```

#### 4d: Reflecting on Differences in Predetermined Characteristics  

From the linear models above we can see that none of the differences between the control and treatment groups across any of the predetermined characteristics are statistically significant because all of their 95% confidence intervals include 0. Practically, we can also assess that none of the differences reported are meaningful in a real world sense. The difference in teacher experience is less that a year, the difference in the fraction of masters degree holders and the difference in the number of Black and White teachers are all less than 1 percentage point. Based on these determinations, we can say that the STAR experiment was successful in randomly assigning teachers to the control and treatment groups. 

## Question 5: Evaluating the Experiment

#### 5a: Multivariate Regression
```{r multi}
model_multi <- lm(sat_index ~ small + factor(school_id), data = star_teachers)
coeftest(model_multi, vcov = vcovHC(model_multi, type="HC1"))
```

#### 5b: Estimated Effect
```{r estimate}
itt_effect <- 0.163630

bounds_multi <- c(0.16363 - 1.96*0.042549, 
                  0.16363 + 1.96*0.042549)

cat("The estimated effect of small class sizes on test scores is", itt_effect, "
    and because the 95% confidence interval does not include zero, we can say 
    this finding is statistically significant.")
```

#### 5c: Bar Chart
```{r bar chart}
# Calculate control group mean score
teachers_ctrl_mean <- mean(star_teachers$sat_index[star_teachers$small == 0])

# Construct dataframe to plot
star_plot_data <- star_teachers %>%
  group_by(small) %>%
  summarise(small = mean(small), 
            score = ifelse(small == 0, 
                           teachers_ctrl_mean, 
                           teachers_ctrl_mean+itt_effect)) %>%
  select(small, score) %>%
  rename (group = small) %>%
    mutate(ub = ifelse(group == 1, bounds_multi[2], NA),
           lb = ifelse(group == 1, bounds_multi[1], NA)) %>%
  mutate(group = ifelse(group == 1, "Treatment", "Control"))

ggplot(data=star_plot_data, aes(x=group, y=score, fill=group)) +
  geom_bar(stat="identity", show.legend = FALSE, width=.6) +
  geom_errorbar(aes(ymin=lb, ymax=ub), 
                width=.1, size = 0.7, color="black") +
  scale_fill_manual(values=c("red", "blue")) +
  labs(y = "SAT Index Score", x = "")
```










