---
title: "HKS SUP-135 Lab 1: Introductory Statistical Concepts and Statistical Computing"
author: "Matt Khinda"
date: "2/3/2023"
output: pdf_document
---

```{r setup, include=FALSE}
# Settings for PDF output
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_chunk$set(fig.width=12, fig.height=6, fig.align = "center") 

# Require packages and install and load if not already
if (!require(haven)) install.packages("haven"); library(haven)
if (!require(tidyverse)) install.packages("tidyverse"); library(tidyverse)
if (!require(ggplot2)) install.packages("ggplot2"); library(ggplot2)
if (!require(here)) install.packages("here"); library(here)
```

```{r load data, include=FALSE}
# Clear workspace
rm(list=ls())
cat('\014')

# Load dataset found in the folder "datasets" 
nlsy <- read_dta(here("datasets", "nlsy97.dta"))
```

## Question 1: Histogram

```{r hist, echo=FALSE}
inc_hist <- ggplot(nlsy) +
  geom_histogram(aes(x = kid_income), bins = 50)

inc_hist
```

## Question 2: Mean

```{r mean, echo=FALSE}
inc_mean <- mean(nlsy$kid_income, na.rm = TRUE)
cat("The mean income for the sample is $", inc_mean, sep="")
```

## Question 3: Conditional Variables

#### 3a: Below the mean
```{r cond 1, echo=FALSE}
nlsy$below_mean <- ifelse(nlsy$kid_income > inc_mean, 0, 1)
```

#### 3b: Percent below mean
```{r cond 2, echo=FALSE}
pct_below_mean <- mean(nlsy$below_mean)*100
cat("The percent of children below the mean income is ", pct_below_mean, "%", sep = "")
```

#### 3c: Why is it not 50%?
```{r cond 3, echo=FALSE}
cat("Because the incomes (shown in the histogram above) are not evenly distributed.")
```

## Question 4: Median

```{r med, echo=FALSE}
inc_med <- median(nlsy$kid_income, na.rm = TRUE)
cat("The median income is $", inc_med, sep = "")
```

## Question 5: Standard Deviation

```{r sd, echo=FALSE}
inc_sd <- sd(nlsy$kid_income)
cat("One stanard deviation is equal to $", inc_sd, sep = "")
```

## Question 6: Wihtin 1 or 2 Standard Deviations

```{r within sd, echo=FALSE}
nlsy$sd1 <- ifelse(abs(nlsy$kid_income - inc_mean) < inc_sd, 1, 0)
nlsy$sd2 <- ifelse(abs(nlsy$kid_income - inc_mean) < inc_sd*2, 1, 0)

cat("The percent of children within one standard deviation is ", mean(nlsy$sd1)*100, "%. ", sep = "")
cat("The percent of children within two standard deviations is ", mean(nlsy$sd2)*100, "%.", sep = "")
```

## Question 7: Percentile Ranks

#### 7a: Rank incomes
```{r pct ranks, echo=FALSE}
nlsy$kid_inc_rank <- rank(nlsy$kid_income, na.last = TRUE)
```

#### 7b: Sort by rank
```{r sort ranks, echo=FALSE}
nlsy[order(nlsy$kid_inc_rank),]
```
#### 7c: Normalize rank
```{r normal ranks, echo=FALSE}
# Define pct_rank function
pct_rank <- function(x){
  #Catch NAs
  r <- ifelse(is.na(x),NA, rank(x,ties.method = "average"))
  #return percentile rank
  100*r/max(r,na.rm = TRUE)
}

# Overwrite kid_inc_rank with new percentile ranks
nlsy$kid_inc_rank <- pct_rank(nlsy$kid_income)
```

#### 7d: Browse the data

## Question 8: Percentile Rank Distribution

#### 8a: Plot percentile rank distribution
```{r rank hist, echo=FALSE}
pctrank_hist <- ggplot(nlsy) +
  geom_histogram(aes(x = kid_inc_rank), bins = 50)

pctrank_hist
```

#### 8b: Validate percentile rank mean and median
```{r rank mean_med, echo=FALSE}
inc_pctrnk_mean <- mean(nlsy$kid_inc_rank, na.rm = TRUE)
inc_pctrnk_med <- median(nlsy$kid_inc_rank, na.rm = TRUE)

cat("The mean percentile rank is", inc_pctrnk_mean, " while the median percentile rank is", inc_pctrnk_med)
```

## Question 9: Relationships

#### Linear correlation  
```{r linear, message = FALSE, echo=FALSE}
linear_rel_plot <- ggplot(nlsy,aes(x = child_education, y = kid_income)) +
  stat_smooth(method = "lm", se = FALSE) +
  stat_summary_bin(fun="mean", bins=10, size=2, geom="point")

linear_rel_plot
```

#### Non-linear correlation  
```{r nonlinear, message = FALSE, echo=FALSE}
nonlinear_rel_plot <- ggplot(nlsy,aes(x = mother_education, y = kid_income)) +
  stat_smooth(method = "lm", se = FALSE) +
  stat_summary_bin(fun="mean", bins=10, size=2, geom="point")

nonlinear_rel_plot
```

## Question 10: Randomization  

#### 10a: Generate and assign random values
```{r rand assign, echo=FALSE}
# Set random seed based on HUID number
HUID <- 41531460
set.seed(HUID)

# Generate list of random values between 0 and 1
random_values <- runif(nrow(nlsy))

# Assign random values to rows in dataframe
nlsy$rand_val <- random_values[nlsy$id_num]
```

#### 10b: Determine treatment group status
```{r tgroup, echo=FALSE}
nlsy$treatment_group <- ifelse(nlsy$rand_val > .5, 1,0)

nlsy_treatment_group <- filter(nlsy, nlsy$treatment_group == 1)
nlsy_control_group <- filter(nlsy, nlsy$treatment_group == 0)

cat("There are", nrow(nlsy_treatment_group), "observations in the treatment group and", nrow(nlsy_control_group), "observations in the control group.")
```

#### 10c: Treatment Group 
```{r tgroup summary, echo=FALSE}
nlsy_treatment_group %>% summarise_all(c("mean","sd"))
```

#### Control Group
```{r cgroup summary, echo=FALSE}
nlsy_control_group %>% summarise_all(c("mean","sd"))
```

#### 10d: Google form submisson
[submitted]

#### 10e: What is the purpose of random assigment in an experiment?
Random assignment seeks to reduce or eliminate selection bias either on the part of participants (in the case of opt-in trials) or the researchers (in the case of researcher selection). For this reason, I would prefer to use random assignment to best achieve comparability.
