---
title: 'HKS SUP-135 Lab 2: Measuring Upward Mobility Using the National Longitudinal
  Survey'
author: "Matt Khinda"
date: "2/10/2023"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# Settings for PDF output
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_chunk$set(fig.width=12, fig.height=6, fig.align = "center") 

# Require packages and install and load if not already
if (!require(haven)) install.packages("haven"); library(haven)
if (!require(tidyverse)) install.packages("tidyverse"); library(tidyverse)
if (!require(ggplot2)) install.packages("ggplot2"); library(ggplot2)
if (!require(statar)) install.packages("statar"); library(statar)
if (!require(here)) install.packages("here"); library(here)

# Clear workspace
rm(list=ls())
cat('\014')
```

```{r load data}
# Load dataset found in the folder "datasets" 
nlsy <- read_dta(here("datasets", "nlsy97.dta"))
```
## Question 1: Subsetting Data

#### 1a: Fraction of Children Whose Mothers have a High School Education or Less Receiving College Degree

```{r mother edu}
# Subset full dataset for children whose mothers have high school education or less
momEdu_under12 <- subset(nlsy, nlsy$mother_education <= 12)
# Calculate fraction of children who have a college degree in this subset 
child_ed_mean <- mean(momEdu_under12$child_college, na.rm = T)

cat("The fraction of children whose mothers have a high school education 
    or less that receive a college degree is ",child_ed_mean*100, "%", sep="")
```

#### 1b: Judgement 

Given the small sample size of our data set, I would conclude that my result of 18.18% is fairly close to Chetty et. al's finding of 20.9%. 


## Question 2: Percentile Ranks

#### 2a: Calculate Child Income Percentile Rank

```{r pct rank kid}
# Define pct_rank function
pct_rank <- function(x){
  #Catch NAs
  r <- ifelse(is.na(x),NA, rank(x,ties.method = "average"))
  #return percentile rank
  100*r/max(r,na.rm = TRUE)
}

# Calculate kid_inc_rank 
nlsy$kid_inc_rank <- pct_rank(nlsy$kid_income)
```

#### 2b: Calculate Parent Income Percentile Rank

```{r pct rank parent}
# Calculate parent_inc_rank 
nlsy$parent_inc_rank <- pct_rank(nlsy$parent_inc)
```


## Question 3: Visualizing

#### 3a: Individual-Level Scatter Plot

```{r scatter_1, message = FALSE}
ggplot(nlsy, aes(x=parent_inc_rank, y=kid_inc_rank)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = 'lm', se = F)
```

#### 3b: Binned Scatter Plot

```{r scatter_2, message = FALSE}
ggplot(nlsy, aes(x=parent_inc_rank, y=kid_inc_rank)) +
  geom_smooth(method = 'lm', se = F) +
  stat_binmean(n = 20, geom = "point")
```
By comparing the two plots above, we can see that the binned scatter plot provides a much clearer picture of the linear relationship between parent's income rank and the child's income rank at adulthood. 



## Question 4: Linear Regression

```{r lm_1}
model_1 <- lm(kid_inc_rank ~ parent_inc_rank, data = nlsy)
summary(model_1)
```

## Question 5: Measures of Upward Mobility

#### 5a: Absolute Mobility at 25th Percentile

```{r measure 1}
absMobility25 <- 31.41826 + .37279*25
cat("The predicted income rank for someone born to parents at the 
    25th percentile rank is", absMobility25)
```

#### 5b: Relative Mobility

```{r measure 2}
relMobility <- (31.41826 + .37279*100)-(31.41826 + .37279*0)
cat("The relative mobility is", relMobility)
```

#### 5c: Horatio Alger Measure

```{r measure 3}
nlsy$top_20 <- ifelse(nlsy$kid_inc_rank > 80, 1,0)
horatioAlger <- with(subset(nlsy, parent_inc_rank < 20), mean(top_20, na.rm = T))
cat("The probability that a child born to parents in the bottom income quintile 
    reaches the top income quintile is ",horatioAlger*100, "%", sep="")
```


#### 5d: Absolute Mobility

```{r measure 4}
#set inflation adjustment factor
infl_adj <- 1.4767

nlsy$kid_inc_greater <- ifelse(nlsy$kid_income > infl_adj*nlsy$parent_inc, 1, 0)
absMobility <- mean(nlsy$kid_inc_greater)
cat("The fraction of children earning more than their parents in 
    inflation-adjusted dollars is ",absMobility*100, "%", sep="")

```


## Question 6: Racial Disparities in Measures of Upward Mobility

#### 6a: Absolute Mobility at 25th Percentile

```{r disparity_measure 1}
#subset data by race and gender
black_men <- subset(nlsy, female == 0 & black ==1)
white_men <- subset(nlsy, female == 0 & white ==1)

black_men_model <- lm(kid_inc_rank ~ parent_inc_rank, data = black_men)
white_men_model <- lm(kid_inc_rank ~ parent_inc_rank, data = white_men)
summary(black_men_model)
summary(white_men_model)

black_men_absMobility25 <- 25.70348 + 0.29432*25
white_men_absMobility25 <- 40.11022 + 0.26692*25

cat("The predicted income rank for Black men born to parents at the 
    25th percentile rank is", black_men_absMobility25, "while for white men it is", white_men_absMobility25)
```

#### 6b: Relative Mobility

```{r disparity_measure 2}
black_men_relMobility <- (25.70348 + 0.29432*100)-(25.70348 + 0.29432*0)
white_men_relMobility <- (40.11022 + 0.26692*100)-(40.11022 + 0.26692*0)

cat("The relative mobility measure for Black men is", black_men_relMobility, 
    "while for white men it is", white_men_relMobility)
```

#### 6c: Horatio Alger Measure

```{r disparity_measure 3}
black_men_horatioAlger <- with(subset(black_men, parent_inc_rank < 20), mean(top_20, na.rm = T))
white_men_horatioAlger <- with(subset(white_men, parent_inc_rank < 20), mean(top_20, na.rm = T))

cat("The probability that a Black man born to parents in the bottom income 
    quintile reaches the top income quintile is ",black_men_horatioAlger*100, 
    "% while for a white man it is ", white_men_horatioAlger*100, "%",sep="")
```


#### 6d: Absolute Mobility

```{r disparity_measure 4}
black_men_absMobility <- mean(black_men$kid_inc_greater)
white_men_absMobility <- mean(white_men$kid_inc_greater)

cat("The fraction of Black men earning more than their parents in 
    inflation-adjusted dollars is ",black_men_absMobility*100, "% while 
    the fraction of white men is ", white_men_absMobility*100, "%",sep="")
```

Of the four measures of mobility presented above, three of them (absolute mobility at the 25th percentile, relative mobility, and the Horatio Alger measure) show a clear disparity in outcomes for white and Black men. However, because absolute mobility is nearly the same for white and Black men it would be misleading to make a determination on that statistic alone. Therefore it does matter which measure you use to determine if there is a racial disparity. 


## Question 7: Regression vs Arithmetic Mean

#### 7a: Arithmetic Mean

```{r mean}
# Subset full dataframe for parent income rank between 20 and 30
subset_btwn_20_30 <- subset(nlsy, parent_inc_rank > 20 & parent_inc_rank < 30)
# Calculate mean child income rank based on that subset
arithmetic_mean_20_30 <- mean(subset_btwn_20_30$kid_inc_rank)
cat("The average (mean) income rank for someone born to parents at the 
    25th percentile rank is", arithmetic_mean_20_30)
```

#### 7b: Random Sample — Linear Regression 

```{r rand measure 1}
# Set random seed based on HUID number
HUID <- 41531460
set.seed(HUID)

# Generate list of random values between 0 and 1
random_values <- runif(nrow(nlsy))

# Assign random values to rows in dataframe
nlsy$rand_val <- random_values[nlsy$id_num]

# Create new dataframe with top 50 highest random values
nlsy_randSubset <- nlsy %>%
  arrange(desc(rand_val)) %>%
  slice(1:50)

# Run linear regression on subset
randSubset_model <- lm(kid_inc_rank ~ parent_inc_rank, data = nlsy_randSubset)
summary(randSubset_model)

# Calculate absolute mobility at 25th percentile
randSubset_absMobility25 <- 33.1733 + .4604*25
cat("The predicted income rank for someone born to parents at the 
    25th percentile rank in this random sample (n=", nrow(nlsy_randSubset),
    ") is ", randSubset_absMobility25, sep = "")

```

#### 7c: Random Sample — Arithmetic Mean

```{r rand measure 2}
randSubset_btwn_20_30 <- subset(nlsy_randSubset, parent_inc_rank > 20 & parent_inc_rank < 30)
randSubset_arithmetic_mean_20_30 <- mean(randSubset_btwn_20_30$kid_inc_rank)
cat("The average (mean) income rank for someone born to parents at the 
    25th percentile rank in this random sample (n=",nrow(randSubset_btwn_20_30),
    ") is ", randSubset_arithmetic_mean_20_30, sep ="")

```

#### 7d: Random Sample — Comparing Measures

As shown in the calculations above on the random sample, the predicted income rank based on the linear regression is much closer than the arithmetic mean to the full population estimate of 41.3 calculated by Chetty et al. 




