---
title: 'HKS SUP-135 Lab 9: The Long-Run Causal Effects of HOLC Redlining'
author: "Matt Khinda"
date: "4/21/2023"
output: pdf_document
---

```{r setup, include=FALSE}
# Settings for PDF output
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 
knitr::opts_chunk$set(fig.width=12, fig.height=5.5, fig.align = "center") 

# Require packages and install and load if not already
if (!require(haven)) install.packages("haven"); library(haven)
if (!require(tidyverse)) install.packages("tidyverse"); library(tidyverse)
if (!require(ggplot2)) install.packages("ggplot2"); library(ggplot2)
if (!require(statar)) install.packages("statar"); library(statar)
if (!require(here)) install.packages("here"); library(here)
if (!require(rpart)) install.packages("rpart"); library(rpart)
if (!require(lmtest)) install.packages("lmtest"); library(lmtest)
if (!require(rdrobust)) install.packages("rdrobust"); library(rdrobust)
if (!require(sandwich)) install.packages("sandwich"); library(sandwich)

# Clear workspace 
rm(list=ls())
cat('\014')
```

```{r load data, include=FALSE}
holc <- read_dta(here("datasets", "holc.dta"))
```

```{r data setup}
# determine distance from threshold
holc$dist_from_cut <- holc$pop_1930 - 40000

# subset data
holc1930 <- subset(holc, year == 1930)
```

## Question 1: Graphical Regression Discontinuity Analysis

#### 1a: HOLC Maps and Population Size
```{r bin_scatter_1}
# Plot discontinuity of HOLC mapping
rdplot(y = holc1930$holc_map, 
       x = holc1930$dist_from_cut,
       c = 0, 
       p = 1,
       nbins = 20,
       binselect = "es",
       x.label = "Distance from 40,000 Population Threshold in 1930",
       y.label = "Likelihood of HOLC Mapping",
       title = "Discontinuity of HOLC Mapping and Population Size"
       )
```

#### 1b: Predetermined Characteristics
```{r bin_scatter_2}
# Plot discontinuity of ownership
rdplot(y = holc1930$ownhome, 
       x = holc1930$dist_from_cut,
       c = 0, 
       p = 1,
       nbins = 20,
       binselect = "es",
       x.label = "Distance from 40,000 Population Threshold in 1930",
       y.label = "Homeownership Rate in 1930",
       title = "Smoothness Test: Homeownership"
       )

# Plot discontinuity of foreign born pop
rdplot(y = holc1930$foreign_born, 
       x = holc1930$dist_from_cut,
       c = 0, 
       p = 1,
       nbins = 20,
       binselect = "es",
       x.label = "Distance from 40,000 Population Threshold in 1930",
       y.label = "Share Foreign Born in 1930",
       title = "Smoothness Test: Foreign Born Population"
       )

# Plot discontinuity of rent
rdplot(y = holc1930$rent, 
       x = holc1930$dist_from_cut,
       c = 0, 
       p = 1,
       nbins = 20,
       binselect = "es",
       x.label = "Distance from 40,000 Population Threshold in 1930",
       y.label = "Mean Rent in 1930",
       title = "Smoothness Test: Mean Rent"
       )
```

#### 1c: Regression Discontinuity Design Validity

Driven by outliers 




#### 1d: Homeownership and Population Size
```{r bin_scatter_3}
# subset data for specific date range
holc_1940_2010 <- subset(holc, year >= 1940 & year <= 2010)

# Plot discontinuity of ownership
rdplot(y = holc_1940_2010$ownhome, 
       x = holc_1940_2010$dist_from_cut,
       c = 0, 
       p = 1,
       nbins = 20,
       binselect = "es",
       x.label = "Distance from 40,000 Population Threshold in 1930",
       y.label = "Homeownership Rate between 1940 and 2010",
       title = "Smoothness Test: Homeownership"
       )

```

## Question 2: Regression Discontinuity Calculations

```{r rd_calcs}
# Indicator for threshold
holc$above_threshold <- ifelse(holc$dist_from_cut >= 0, 1, 0)

# Create interaction term
holc$interaction_term <- (holc$dist_from_cut * holc$above_threshold)

# calculate rdd 
rdd <- lm(ownhome ~ above_threshold + 
            dist_from_cut + 
            interaction_term, 
          data = holc_1940_2010)

coeftest(rdd, vcovCL(rdd, cluster = holc_1940_2010$city_id))
```

## Question 3: Graphical Difference-in-Differences Analysis

```{r diff_plots}
holc$treated <- holc$above_threshold

ggplot(holc, aes(x = year, 
                 y = ownhome, 
                 color = factor(treated, 
                                labels = c("1930 pop < 40k (Treatment)", 
                                           "1930 pop > 40k (Control)")),
                 shape = factor(treated, 
                                labels = c("1930 pop < 40k (Treatment)", 
                                           "1930 pop > 40k (Control)")))) +
      geom_vline(xintercept = 1935) +
      stat_summary(fun = "mean", geom = "point") +
      stat_summary(fun = "mean", geom = "line") + 
      labs(x = "Year", y = "Home Onwership Rate", shape = "", color = "") + 
      theme(legend.position = "bottom") 
```

## Question 4: Conditional Means

```{r cond_means}
# Calculate difference for treatment group pre and post
Y_treat_pre <- with(subset(holc, treated ==1 & year >= 1910 & year <= 1930), 
                    mean(ownhome, na.rm = T))
Y_treat_post <- with(subset(holc, treated ==1 & year >= 1940 & year <= 1960), 
                     mean(ownhome, na.rm = T))
Y_treat_diff <- Y_treat_post - Y_treat_pre 

# Calculate difference for control group pre and post
Y_control_pre <- with(subset(holc, treated ==0 & year >= 1910 & year <= 1930), 
                    mean(ownhome, na.rm = T))
Y_control_post <- with(subset(holc, treated ==0 & year >= 1940 & year <= 1960), 
                     mean(ownhome, na.rm = T))
Y_control_diff <- Y_control_post - Y_control_pre 

# Calculate difference in differences
Y_diffdiff <- Y_treat_diff - Y_control_diff
```

## Question 5: Difference-in-Differences Calculations

```{r diff_calcs}
# Subset data
holc_1910_1960 <- subset(holc, year >= 1910 & year <= 1960)

# Assign indicator variable for post 1940
holc_1910_1960$post <- ifelse(holc_1910_1960$year >= 1940, 1, 0)

# Define interaction term
holc_1910_1960$dd_interaction_term <- (holc_1910_1960$above_threshold*holc_1910_1960$post)

# Create diff in diff regression
dd_lm <- lm(ownhome ~ dd_interaction_term + post + above_threshold, data = holc_1910_1960)
coeftest(dd_lm, vcovCL(dd_lm, cluster = holc_1910_1960$city_id))
```

## Question 6: Fixed-Effects Regression

```{r fixed_effects}

```

## Question 7: Causal Effect of HOLC Redlining on Homeownership







